---
title: "Assignment1-Pathak-Cause Of Death 2014 2019"
author: "Rohan Pathak"
date: "2023-03-07"
output: html_document
---
# Introduction
This dataset contains the number of deaths which occured in the United States of America due to broad selected causes. The data is divided into months for each year between 2014 to 2019. 

Dataset Name- Monthly Counts of Deaths by Select Causes, 2014-2019 

Dataset Source- U.S. Department of Health & Human Services

Dataset Link- https://catalog.data.gov/dataset/monthly-counts-of-deaths-by-select-causes-2014-2019-da9df

# Variable Names
```{r}
#Jurisdiction of Occurrence- Where deaths occured
#Year- Year of death
#Month- Month of death
#All Cause- Number of deaths by all causes
#Natural Cause- Number of deaths by the cause
#Septicemia- Number of deaths by the cause
#Malignant Neoplasms- Number of deaths by the cause
#Diabetes Mellitus- Number of deaths by the cause
#Alzheimer Disease- Number of deaths by the cause
#Influenza and Pneumonia- Number of deaths by the cause
#Chronic Lower Respiratory Diseases- Number of deaths by the cause
#Other Diseases of Respiratory System- Number of deaths by the cause
#Nephritis, Nephrotic Syndrome, and Nephrosis- Number of deaths by the cause
#Symptoms, Signs, and Abnormal Clinical and Laboratory Findings, Not Elsewhere Classified- Number of deaths by the cause
#Diseases of Heart- Number of deaths by the cause
#Cerebrovascular Diseases- Number of deaths by the cause
#Accidents (Unintentional Injuries)- Number of deaths by the cause
#Motor Vehicle Accidents- Number of deaths by the cause
#Intentional Self-Harm (Suicide)- Number of deaths by the cause
#Assault (Homicide)- Number of deaths by the cause
#Drug Overdose- Number of deaths by the cause
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Importing libraries
```{r message=FALSE,warning=FALSE}
#install.packages("ggplotify")
library(kableExtra)
library(tidyverse)
library(zoo)
library(ggplot2)
library(ggthemes)
library(stats)
#install.packages("seasonalityPlot")
library(forecast)
#install.packages('hrbrthemes')
library(hrbrthemes)
library(cowplot)
```

# Loading the messy data to R
The data in question is. The data has been picked from. The variables are
```{r message=FALSE,warning=FALSE}
df=read_csv(file="Assignment1-Pathak-Cause Of Death 2014 2019.csv")
summary(df)
```

The data has multiple columns for different causes of death, which is the same kind of observation, thus they should instead be placed in a single column according to the rules of tidy data.
```{r message=FALSE,warning=FALSE}
df[1:10, 1:5] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)
```

# Data Tidying-

Tidying the data by pivot longer on the different columns capturing cause of death into a single column called "Cause of Death" with the values moved to a new column "Cases"
```{r message=FALSE,warning=FALSE}

tid_df<-df %>%
  pivot_longer(c("All Cause", "Natural Cause", "Septicemia",
                "Malignant Neoplasms", "Diabetes Mellitus", 
                "Alzheimer Disease", "Influenza and Pneumonia", 
                "Chronic Lower Respiratory Diseases", 
                "Other Diseases of Respiratory System", 
                "Nephritis, Nephrotic Syndrome, and Nephrosis", 
                "Symptoms, Signs, and Abnormal Clinical and Laboratory Findings, Not Elsewhere Classified", "Diseases of Heart", 
                "Cerebrovascular Diseases", "Accidents (Unintentional Injuries)",
                "Motor Vehicle Accidents", "Intentional Self-Harm (Suicide)",
                "Assault (Homicide)", "Drug Overdose"), 
               names_to = "Cause_of_Death", values_to = "Cases")

```

Now the dataset is following the three main rules of tidy dataset- 
Each variable must have its own column. - Fixed by pivot_longer
Each observation must have its own row. - Fixed by pivot_longer
Each value must have its own cell.
```{r message=FALSE,warning=FALSE}
tid_df[1:10, 1:5] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)
```
# Data Wrangling-

Now doing Exploratory Data Analysis, we see there are no missing values, the column Jurisdiction of Occurrence" has only one value, so it can be dropped.
```{r message=FALSE,warning=FALSE}
glimpse(tid_df)
tid_df<-tid_df[,-1]
tid_df[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)
```

We can also combine the Year and Month columns into a single date using unite as described below.
```{r message=FALSE,warning=FALSE}
tid_df2<-tid_df
tid_df2 %>%
  unite(Year, Month, Year, sep=" ")
```

Instead we will use zoo yearmon function to get a date object. Removing the Month and Year column, then moving the newly created Date column to the beginning.
```{r message=FALSE,warning=FALSE}
tid_df$Date <- as.yearmon(paste(tid_df$Year, tid_df$Month), "%Y %m")
tidy_df <- tid_df %>%
  select(Date, everything())
tidy_df<-tidy_df[,-2]
tidy_df<-tidy_df[,-2]
class(tidy_df$Date)

tidy_df[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)
```

Creating a dataframe to add day and arrange date to %Y-%m-%d format and sort by date 
```{r message=FALSE,warning=FALSE}
df3<-tidy_df
df3$Date <- as.Date(df3$Date)
df3 %>% arrange(mdy(df3$Date))
df3<-df3[order(as.Date(df3$Date, format="%Y-%m-%d")),]
df3[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)
```
# Data Visualization with inferences-

Creating plot with just the data for "All Cause" to see the outlier months with most deaths
```{r message=FALSE,warning=FALSE}
df2<-df3[df3$Cause_of_Death == 'All Cause', ]
df2$Date <- as.Date(df2$Date)
pl1<-ggplot(data = df2,
       mapping = aes(x = factor(Date),
                     y = Cases)) +
  geom_point() +
  geom_smooth() +
  theme_economist() +
  scale_colour_economist()+
  labs(title = "Deaths by Cause",
       x = "Date",
       y = "Cases") +
  theme(axis.text.x=element_blank(),
      )

pl2<-ggplot(df2, aes(x = Date, y = Cases, color = Cause_of_Death,
                group = Cause_of_Death)) +
  geom_line(size = 1) +
  scale_x_date(date_labels = "%b-%y", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Date", y = "Number of Cases", color = "Cause of Death") +
  ggtitle("Over Time Cases") +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.position = "bottom"
        ) + 
  theme_economist() +
  scale_colour_economist()+
  theme(axis.text.x=element_blank(),)

plot_grid(pl1, pl2, labels = "AUTO")
```


As we can see from the above plots, the overall trend of the total number of deaths is increasing over the years, with certain months being outliers. Like in the above plots, the months of January and December over the years are the outliers. This proves the conventional wisdom that winter months have the most deaths

Since the majority of deaths are due to natural causes, we drop the rows
for them to see the significant reasons for other causes of deaths
```{r message=FALSE,warning=FALSE}
df_drop<-df3[is.element(df3$Cause_of_Death, c("Septicemia",
                                              "Malignant Neoplasms", "Diabetes Mellitus", 
                                              "Alzheimer Disease", "Influenza and Pneumonia", 
                                              "Chronic Lower Respiratory Diseases", 
                                              "Other Diseases of Respiratory System", 
                                              "Nephritis, Nephrotic Syndrome, and Nephrosis", 
                                              "Symptoms, Signs, and Abnormal Clinical and Laboratory Findings, Not Elsewhere Classified",
                                              "Diseases of Heart", 
                                              "Cerebrovascular Diseases", "Accidents (Unintentional Injuries)",
                                              "Motor Vehicle Accidents", "Intentional Self-Harm (Suicide)",
                                              "Assault (Homicide)", "Drug Overdose")),]

df_drop[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)

ggplot(df_drop, aes(x = Date, y = Cases, color = Cause_of_Death, group = Cause_of_Death)) +
  geom_line(size = 1) +
  scale_x_date(date_labels = "%b-%y", date_breaks = "1 month") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Date", y = "Number of Cases", color = "Cause of Death") +
  ggtitle("Number of Cases by Cause of Death Over Time") +
  theme(plot.title = element_text(size = 20, hjust = 0.5),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.position = "bottom") +
  theme(axis.text.x=element_blank(),)
```


The initial observation about winter months holds true, thus the data is seasonal in nature. It is interesting to see that the trend is valid for most of the causes of deaths, with the sharpest jumps in diseases of heart. We also see that deaths due to Diseases of Heart and Malignant Neoplasm far outweighs other reasons.



Now to find the change in the top contributor to deaths over the years we decide to look at the data on a year level. We group by year finding average and then drop all rows except for 2014 and 2019, to see overall change.
```{r message=FALSE,warning=FALSE}
df_drop3<-df_drop %>%
  mutate(day = format(Date, "%d"),month = format(Date, "%m"),
         year = format(Date, "%Y")) %>%
  group_by(year,Cause_of_Death) %>%
  summarise(total = sum(Cases))

df_drop3[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead = T)

ggplot(df_drop3, aes(x=Cause_of_Death, y=total, fill=Cause_of_Death)) +
  geom_bar(stat="identity") +
  xlab("Cause of Death") +
  ylab("Number of Deaths") +
  ggtitle("Number of Deaths by Cause between 2014 to 2019")  +
  theme(axis.text.x=element_blank(),)
```


From the above graph we can see that after Diseases of Heart and Malignant Neoplasms, the top three causes are Accidents, Cerebrovascular, and Respiratory diseases

Now to find the overall changes in the top causes over the 5 year period, we filter the dataset to only include data for 2014 and 2019
```{r message=FALSE,warning=FALSE}
deaths_2014 <- df_drop3[df_drop3$year == 2014,]
deaths_2019 <- df_drop3[df_drop3$year == 2019,]
```

Sort the data by total number of deaths in descending order
```{r message=FALSE,warning=FALSE}
deaths_2014 <- deaths_2014[order(-deaths_2014$total),]
deaths_2019 <- deaths_2019[order(-deaths_2019$total),]

```

Create a bar plot of the top 10 causes of death in 2014 and 2019
```{r message=FALSE,warning=FALSE}
gplot1<-ggplot(data = head(deaths_2014, 10), aes(x = reorder(Cause_of_Death, total), y = total)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  ggtitle("Top 10 Causes of Death in 2014") +
  xlab("Cause of Death") +
  ylab("Total Deaths") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
gplot2<-ggplot(data = head(deaths_2019, 10), aes(x = reorder(Cause_of_Death, total), y = total)) +
  geom_bar(stat = "identity", fill = "red") +
  ggtitle("Top 10 Causes of Death in 2019") +
  xlab("Cause of Death") +
  ylab("Total Deaths") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

plot_grid(gplot1, gplot2, labels = "AUTO")
```


Accidents seem to have overtaken the number of deaths due to Respiratory diseases in 2019. Another observation, the number of deaths due to Heart disease has markedly increased compared to due to Neoplasms which seems to be close to the 2014 number.

Now to see overall trends in changes for Cause of death. We group data to year
```{r message=FALSE,warning=FALSE}
deaths_grouped <- aggregate(total ~ Cause_of_Death + year, data = df_drop3, sum)

deaths_grouped[1:10,] %>%
  kbl() %>%
  kable_paper("hover", full_width = F, fixed_thead=T)
```

Create a line plot of the total deaths for each cause of death over the years 2014-2015
```{r message=FALSE,warning=FALSE}
ggplot(data = deaths_grouped, aes(x = year, y = total, group = Cause_of_Death, color = Cause_of_Death)) +
  geom_line() +
  ggtitle("Total Deaths by Cause of Death (2014-2019)") +
  xlab("Year") +
  ylab("Total Deaths") +
  theme_ipsum() + # Arial Narrow
  scale_fill_ipsum() +
  theme(legend.position = "top")
```


We see that the observations from the last plot on a moving trend level. The deaths by heart disease are increasing, while melanoma death numbers are stable. Also the deaths due to accidents overtook deaths by respiratory disease in mid 2015.
